---
title: "Print complex `{xaringan}` slides"
author: "Bernhard Bieri"
date: "`r Sys.Date()`"

output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Print complex `{xaringan}` slides}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r global.options, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.align = 'center',
  fig.width=6,
  fig.height=4
)
```

Printing slides created with the `{iheiddown}` presentation template works
fine in 90% of the time with an internet browser's print function. However,
for more complex layouts, such as revealing slides, one may want greater
control over the printing output. This article, and the function below,
aim at solving this issue.

The function itself is borrowed from Garrick Aden Buie's excellent
[blog](https://www.garrickadenbuie.com/blog/print-xaringan-chromote/) and
leverages `{chromote}`, a headless version of the Chrome web browser to flick
through every slide automatically and print them out one by one.

Note that this function is not included into `{iheiddown}` package as it relies
on the following packages (some of which are not on CRAN yet): `{chromote}`,
`{progress}`, `{jsonlite}`, `{pdftools}`, and `{digest}`.

## The workflow:

1. Open RStudio and the HTML version of the xaringan presentation

2. Copy paste the function below into your session and add it to your global
    environment

3. Run `xaringan_to_pdf()` with the path to your original HTML presentation
    and specify an `output_file` path where you want the PDF version of your
    presentation to be printed.

4. Wait for the function to flick through all the slides and check out the
   result!


```{r PrintFunction, eval=FALSE}
#' Print xaringan slides to PDF
#'
#' Prints xaringan slides to a PDF file, even complicated slides
#' with panelsets or other html widgets or advanced features.
#' Requires a local installation of Chrome as well as the following packages:
#' `{chromote}`, `{progress}`, `{jsonlite}`, `{pdftools}`, and `{digest}`.
#'
#' @param input Path to Rmd or html file of xaringan slides.
#' @param output_file The name of the output file. If using NULL then
#'   the output filename will be based on filename for the input file.
#'   If a filename is provided, a path to the output file can also be provided.
#' @param delay Seconds of delay between advancing to and printing
#'   a new slide.
#' @param include_partial_slides Should partial (continuation) slides be
#'   included in the output? If `FALSE`, the default, only the complete slide
#'   is included in the PDF.
#' @source <https://www.garrickadenbuie.com/blog/print-xaringan-chromote/>

xaringan_to_pdf <- function(
  input,
  output_file = NULL,
  delay = 1,
  include_partial_slides = FALSE
) {
  if (!requireNamespace("chromote", quietly = TRUE)) {
    stop("`chromote` is required: devtools::install_github('rstudio/chromote')")
  }
  required_packages <- c("progress", "jsonlite", "pdftools", "digest", "fs")
  for (pkg in required_packages) {
    if (!requireNamespace(pkg, quietly = TRUE)) {
      stop("`", pkg, "` is required: install.packages('", pkg, "')")
    }
  }
  
  is_url <- grepl("^(ht|f)tp", tolower(input))
  
  if (is.null(output_file)) {
    if (is_url) {
      output_file <- fs::path_ext_set(fs::path_file(input), "pdf")
    } else {
      output_file <- fs::path_ext_set(input, "pdf")
    }
  }
  
  if (!is_url && !grepl("^file://", input)) {
    if (!tolower(fs::path_ext(input)) %in% c("htm", "html")) {
      stop("`input` must be the HTML version of the slides.")
    }
    input <- paste0("file://", fs::path_abs(input))
  }
  
  b <- chromote::ChromoteSession$new()
  on.exit(b$close(), add = TRUE)
  
  b$Page$navigate(input, wait_ = TRUE)
  b$Page$loadEventFired()
  
  has_remark <-
    b$Runtime$evaluate("typeof slideshow !== 'undefined'")$result$value
  if (!has_remark) {
    stop("Input does not appear to be xaringan slides: ", input)
  }
  
  current_slide <- function() {
    x <- b$Runtime$evaluate("slideshow.getCurrentSlideIndex()")$result$value
    as.integer(x) + 1L
  }
  
  slide_is_continuation <- function() {
    b$Runtime$evaluate(
      "document.querySelector('.remark-visible').matches('.has-continuation')"
    )$result$value
  }
  
  hash_current_slide <- function() {
    digest::digest(b$Runtime$evaluate(
      "document.querySelector('.remark-visible').innerHTML"
    )$result$value)
  }
  
  get_ratio <- function() {
    r <- b$Runtime$evaluate('slideshow.getRatio()')$result$value
    r <- lapply(strsplit(r, ":"), as.integer)
    width <- r[[1]][1]
    height <- r[[1]][2]
    page_width <- 8/width * width
    list(
      width = as.integer(908 * width / height),
      height = 681L,
      page = list(width = page_width, height = page_width * height / width)
    )
  }
  
  slide_size <- get_ratio()
  
  expected_slides <- as.integer(
    b$Runtime$evaluate("slideshow.getSlideCount()")$result$value
  )
  
  max_slides <- expected_slides * 4
  
  b$Browser$setWindowBounds(1, bounds = list(
    width = slide_size$width,
    height = slide_size$height
  ))
  
  b$Emulation$setEmulatedMedia("print")
  b$Runtime$evaluate(paste0(
    "let style = document.createElement('style')\n",
    "style.innerText = '@media print { ",
    ".remark-slide-container:not(.remark-visible){ display:none; }",
    if (include_partial_slides) " .has-continuation { display: block }",
    "}'\n",
    "document.head.appendChild(style)"
  ))
  
  pb <- progress::progress_bar$new(
    format = "Slide :slide (:part) [:bar] Eta: :eta",
    total = expected_slides
  )
  
  idx_slide <- current_slide()
  last_hash <- ""
  idx_part <- 0L
  pdf_files <- c()
  for (i in seq_len(max_slides)) {
    if (i > 1) {
      b$Input$dispatchKeyEvent(
        "rawKeyDown",
        windowsVirtualKeyCode = 39,
        code = "ArrowRight",
        key = "ArrowRight",
        wait_ = TRUE
      )
    }
    
    if (current_slide() == idx_slide) {
      step <- 0L
      idx_part <- idx_part + 1L
    } else {
      step <- 1L
      idx_part <- 1L
    }
    idx_slide <- current_slide()
    pb$tick(step, tokens = list(slide = idx_slide, part = idx_part))
    
    if (!isTRUE(include_partial_slides) && slide_is_continuation()) next
    Sys.sleep(delay)
    
    this_hash <- hash_current_slide()
    if (identical(last_hash, this_hash)) break
    last_hash <- this_hash
    
    pdf_file_promise <- b$Page$printToPDF(
      landscape = TRUE,
      printBackground = TRUE,
      paperWidth = 12,
      paperHeight = 9,
      marginTop = 0,
      marginRight = 0,
      marginBottom = 0,
      marginLeft = 0,
      pageRanges = "1",
      preferCSSPageSize = TRUE,
      wait_ = FALSE
    )$then(function(value) {
      filename <- tempfile(fileext = ".pdf")
      writeBin(jsonlite::base64_dec(value$data), filename)
      filename
    })
    pdf_files <- c(pdf_files, b$wait_for(pdf_file_promise))
  }
  
  pdftools::pdf_combine(pdf_files, output = output_file)
  fs::file_delete(pdf_files)
  
  invisible(output_file)
}
```
